Using SMT engine to generate Symbolic Automata\thanks{This work was partially funded by the Associated Team FM4CPS between INRIA and ECNU, Shanghai}

### 2018-01-22 2017-Xudong-SMT/FM-2018-extended.tex
Finished proof-reading -- many comments

\noteSB{$E$ is never updated, so $E = \emptyset$}
Algorithm~\ref{alg1} starts with an open pNet, and builds its set of open
transitions. Its main loop is a classical residual algorithm: starting
from the initial global state, it picks a state in an unexplored set, and
computes all possible OTs, adding their target states in the
unexplored set, until this set is empty. As the potential set of
global states is finite (see \cite{henrio:Forte2016}), this terminates.

\begin{algorithm}[h]
  \caption{Open Automaton Generation}
  \label{alg1}
\begin{algorithmic}[1]
\Require The pNet node P.
\State Initialize sets $U=\mathit{InitState}(P)$, $E=\emptyset$ for unexplored/explored global states, $L=\emptyset$ for the resulting OTs;
\While{!$\mathit{isEmpty}(U)$}
	\State Choose $S$ in $U$;
	\State $\mathit{OT} = \mathit{MakeTransitions}(P, S)$;
        \State Store $\mathit{OT}$ in $L$; 
	\ForAll{$\mathit{OT} \in L$}
        \State Let $T$ be the target global state of $\mathit{OT}$
	\IfThen{($T \not\in U$ \&\& \hl{$T \not\in E$})}{Add $T$ into $U$;}
	\EndFor
\EndWhile
\State Prune the open transitions in $L$ using the SMT solver;
\State \Return $\mathit{OA}=\mathit{Reachable}(\mathit{InitState}(P),L)$;

@@

The inside loop (\emph{MakeTransitions} method) applies recursively
the semantic rules following the structure of the pNet. 
For pLTSs at the leaves, we simply take the pLTS transitions of the
corresponding local state and build the predicate to generate the
open transition. 
On a pNet node we use two methods, combining and
matching, to generate the predicate of the open transitions in a
hierarchical manner.
%
Then, at the root of the pNet, the predicate of each OT is translated
into SMTlib assertions, and checked for satisfiability. Unsatisfiable
OTs are pruned, and the final open automaton is computed as the set of
global states reachable by satisfiable OTs.

\begin{algorithm}
\caption{MakeTransitions()}
  \label{alg2}
\begin{algorithmic}[1]
\Require a pNet node Pn; a global state S.
\State Initialize a list $l$, $L$ for sub-transitions/transitions.
	\For{each Subnet in Pn}
	~~~ $\setminus \setminus$ Recursively applying Tr1 or Tr2 on the sub-nodes.
                \State Store MakeTransitions(Subnet, S) in $l$;
	\EndFor
	\For{each $sv \in SV$}
		\State $\overline{comb} = Combining(l)$;
                 \State $ot = Matching(sv, \overline{comb}, hole, v)$;
                 \State Store $ot$ in $L$; ~~~ $\setminus \setminus$ if Filter(ot) succeeds
	\EndFor
\State \Return $L$;
\end{algorithmic}  
\end{algorithm}

@@

\def\inactive{\{-\}}
\paragraph{Combining:}
The combining method enumerate all the possible behaviours of the
subnets, as all the possible combinations of their open transitions.
Assume that there is a collection of $n$ subnets
$L = {s_1, s_2, ...,  s_n}$; we denote $\overline{ot}_i$ the set of open
transitions of the 
subnet $s_i$ and ``$-$'' means the subnet is not involved;
then the combination $\overline{comb}$, a set of n-tuples, is the
cartesian product of sets :  
$\overline{comb} : (\inactive\cup \overline{ot}_1) \times (\inactive\cup \overline{ot}_2)\times ... \times (\inactive\cup \overline{ot}_n).$

@@

\paragraph{Matching:}
The next step, for each synchronisation vector, and each possible
combination of behaviour of the subnets as generated by the \emph
{Combining} method, consists in building the combined predicate,
and the corresponding open transition.
From a synchronization vector
$sv_k\in SV = (a'_i)^{i\in I} (b'_j)^{j\in J}\rightarrow v'$
and its guard $G_k$, a tuple of open transitions
$C_n\in \overline{comb} = {(v_i)}^{i\in I}$, the hole behaviors $Hole$,
and a fresh variable representing the result action $v$, we build the
predicate: 
%
\begin{multline*}
  \mathit{Matching}(SV_k, C_n, \mathit{Hole}, v) = 
  \\
  \forall {(a'_i)}^{i\in I},
          {(b'_j)}^{j\in J},v'.\
          v_i=a'_i\land \, b_j=b'_j \land v=v' \land G_k
\end{multline*}

\paragraph{Filtering:}
There may be several reasons why such a matching would fail:
\begin{itemize}
  \item if some subnet is marked as inactive in the vector, and the
    chosen combination has an active behaviour at this position,
  \item if some subnet action expression in the vector does not match (by
    pattern-matching) the corresponding behaviour $C_n$,
  \item or if the whole set of active subnet actions in the vector
    cannot be matched (by unification) with the set $\{C_n\}$.
\end {itemize}

Each of these possible filters is more precise than the previous
ones. But even when unification succeeds, it is still possible that
the resulting predicate would be unsatisfiable, because of some
incompatibility involving the guards. In our algorithm, we choose to
apply only the simplest filter inside the \emph{Matching} method. All
the rest will be checked later, using the satisfiability check in the
SMT engine.

@@

In a pLTS, there may be several incoming transitions 
that assign potentially different values to a state variable.
To handle such cases, the algorithm manages for each pLTS state, as a
list of expressions collected from the assignments of each state
variable $v$.
%% The list we used contains triples <$v$, $S$, $AssignRH$> 
%% where $v$ is the variable in pLTS, $S$ is its owner state and
%% $AssignRH$ is a list of expressions over other variables in the right
%% hand side of assignments of $v$.
% As a pragmatic extension to the formal definition, we also manage
% ``global variables'', defined in all states of the pLTS.
For a global state in the open automaton, getting the set of state
variables (which may be used in a transition) will simply be the union
of state variables of the individual pLTS states constituting this
global state (variables of different pLTS are distinct).
%There are some assignment on the variables in each of states of the pLTS (state variables), or just as global states of it (global variables). There is usually not only one assignment for a state variable, miss of them may drop the unsatisfiable results. In order to avoid such mistakes,

@@

Our matching/filtering strategy builds some transitions where
the predicates express incompatible constraints. Even if having an
unsatisfiable (symbolic) transition in the open automaton would not be incorrect,
we want to check them for satisfiability, and reduce the number of
transitions and states in the automaton.
In Fig.~\ref{schema:unsat-ot}, we display an example of an
unsatisfiable open transition 
from the result of our example. It shows the case where
the failure controller performs a ``$fail$'' action, while the timer
executes a ``$resume$''. The chosen synchronization
vector (SV0 from Fig.~\ref{schema:ArchFailure:pNet}) does not match
with these actions, as shown by the predicate
fragment ``resume(true)=start($b_2$)''.

@@

Checking satisfiability requires some symbolic computation
on the action expressions and the predicates, and this 
may depend on the specific theory of the action algebra datatypes. 
The ``Modulo Theory'' part of SMT solvers is important here, so that
the solver can use specific properties of each action algebra.

@@

We check the satisfiability of each open transitions
using the SMT solver Z3. More precisely, we translate
the OT predicate into a SMTlib assertion, in the context provided by
the possible state variable assignments in the source state. In this
section we describe the translation of the algebra presentation, of
assignments, and of the predicates.

From inside our algorithm code, we submit satisfiability requests to
Z3 using its JAVA API. Here for readability, we
show the Z3 code using its SMT-LIB input language.

Figure~\ref{schema:smt-lib} shows the translation of the
transition of Fig.~\ref{schema:unsat-ot}
in the SMTlib syntax.
It contains the declaration
of the BIP action algebra sorts 
and constructors, then the declaration of variables, and finally the
predicate to be checked, encoded as a set of assertions.
Here we display also the proof generated by Z3,
though in practice our tool only needs the brute information ``sat''
or ``unsat''. The result "unsat" in the output is just what we expected.
% as appears in the ``unsat'' proof generated by Z3 in which ``\{= ?x66 ?x70\}'' is shown to be false (see Fig. \ref{schema:smt-lib})

@@

To build the input submitted to Z3 for each OT,
we translate the algebra $presentation$, the predicates and the
variable assignments (the Post part) into Z3 (Java-API) syntax.
% The $presentation$ and the predicates are translated separately.

\paragraph{Translation of Action algebra presentation}
In appendix \ref{appendix:mapsmt} we define formally a condition of
well-formness of an algebra presentation, and its translation into
SMTlib declarations (\texttt{declare-datatypes} and
\texttt{declare-fun}). It ensures that the generated code is correct,
and will not raise runtime errors in the SMT engine. Remark that the
\texttt{declare-datatypes} command include both the Action
constructors from table \ref{table:BIPalgebra}, but also the constant action
names from our example.

\paragraph{Translating assignments into predicate terms}
State variable assignments that were collected during the open
automaton are used when generating the SMTlib asserts (see
appendix~\ref{appendix:translating-assignments}). This appears in
Fig.~\ref{schema:smt-lib} in line 10, that encodes possible values of
the variable $t$ of the timer.

@@

\paragraph{Checking satisfiability}

In App.~\ref{appendix:mapsmt} we formally define the translation of each
open transition, including:
- collect all variables in the transition, and declare them,
- check well-formness and correct typing of expressions
- translate the predicate into a conjunction of assertions
- if present in the source state, translate the state-variable
assignments into a disjunctive assertion.
Here again, this translation ensures that no runtime error will occur
when running the SMT engine.

>>>>

\noteSB{line 8: $E$ is never updated, so $E = \emptyset$ always;\\
line 11: Would it not be better to prune unsatisfiable transitions on-the-fly, \ie inside the loop?}
Algorithm~\ref{alg1} starts with an open pNet, and builds its set of open
transitions. Its main loop is a classical residual algorithm: starting
from the initial global state, it picks a state in an unexplored set, and
computes all possible OTs, adding their target states in the
unexplored set, until this set is empty. As the potential set of
global states is finite (see \cite{henrio:Forte2016}), this terminates.

\begin{algorithm}[h]
  \caption{Open Automaton Generation}
  \label{alg1}
\begin{algorithmic}[1]
\Require The pNet node P.
\State Initialize sets $U=\mathit{InitState}(P)$, $E=\emptyset$ for unexplored/explored global states, $L=\emptyset$ for the resulting OTs;
\While{!$\mathit{isEmpty}(U)$}
	\State Choose $S$ in $U$;
	\State $\mathit{OT} = \mathit{MakeTransitions}(P, S)$;
        \State Store $\mathit{OT}$ in $L$; 
	\ForAll{$\mathit{OT} \in L$}
        \State Let $T$ be the target global state of $\mathit{OT}$
	\IfThen{($T \not\in U$ \&\& \hl{$T \not\in E$})}{Add $T$ into $U$;}
	\EndFor
\EndWhile
\State \hl{Prune the open transitions in $L$ using the SMT solver;}
\State \Return $\mathit{OA}=\mathit{Reachable}(\mathit{InitState}(P),L)$;

@@

The inside loop (\emph{MakeTransitions} method) applies recursively
the semantic rules following the structure of the pNet.
\hl{For pLTSs at the leaves, we simply take the pLTS transitions of the
corresponding local state and build the predicate to generate the
open transition.}
On a pNet node we use two methods, combining and
matching, to generate the predicate of the open transitions in a
hierarchical manner.
%
Then, at the root of the pNet, the predicate of each OT is translated
into SMTlib assertions, and checked for satisfiability. Unsatisfiable
OTs are pruned, and the final open automaton is computed as the set of
global states reachable by satisfiable OTs.

%
\noteInSB{Algorithm~\ref{alg2}, as presented below is not
  well-defined: when we reach the leaves, a pLTS does not have any
  subnets, hence the loop in lines 2--4 is empty and so is the list
  $l$. $\mathit{Combining}(l)$ is then called on an empty list...}

\noteInSB{Line 2: Tr1 and Tr2 are the semantic rules, right?}

\noteInSB{Line 6: Nothing in the assignment $\overline{comb} =
  \mathit{Combining}(l)$ depends on $sv$.  Hence, this should probably
  be done outside of the loop.}

\noteInSB{Line 7: There are two type mismatches here: $ot$ is an open
  transition, whereas $\mathit{Matching}$ is a predicate\mdash one
  cannot assign a predicate to an open transition.  Furthermore, in
  the definition of $\mathit{Matching}$, below, $C_n$ is an open
  transition, whereas $\overline{comb}$ is a {\em set} of open
  transitions, so this call does not type either. There is probably
  another inner loop over all tuples of OTs in $\overline{comb}$; even
  so the open transition has to be defined completely, not just the
  predicate\mdash what is the value of $Post$ for the new transition?
  Maybe $\mathit{Matching}$ should not be a predicate, but an OT?
  Finally, how is $\mathit{hole}$? (I have removed $v$\mdash here and
  in the parameters of the definition of the $\mathit{Matching}$
  predicate, because the text in the paragraph ``Matching'' says that
  $v$ is fresh.)}

\noteInSB{Line 8: When is $\mathit{Filter}(ot)$ called?}

\begin{algorithm}
\caption{MakeTransitions()}
  \label{alg2}
\begin{algorithmic}[1]
\Require a pNet node $Pn$; a global state $S$.
\State Initialize empty list $l$ and set $L$ for sub-transitions/transitions.
	\For{each $\mathit{Subnet}$ in $Pn$}
	~~~ $\setminus \setminus$ Recursively applying \hl{Tr1 or Tr2} on the sub-nodes.
                \State Store $\mathit{MakeTransitions}(Subnet, S)$ in $l$;
	\EndFor
	\For{each $sv \in SV$}
		\State $\overline{comb} = \mathit{Combining}(l)$;
                 \State \hl{$ot = \mathit{Matching}(sv, \overline{\mathit{comb}}, \mathit{hole})$;} % SB: removed the fourth parameter v here
                 \State Store $ot$ in $L$; ~~~ \hl{$\setminus \setminus$ if $\mathit{Filter}(ot)$ succeeds}
	\EndFor
\State \Return $L$;
\end{algorithmic}  
\end{algorithm}

@@

\def\inactive{\{-\}}
\paragraph{Combining.}
The combining method enumerates all the possible behaviours of the
subnets, as all the possible combinations of their open transitions.
Assume that there is a collection of $n$ subnets.\noteSB{These variables are not used and conflict with the returned set $L$ in Alg.~\ref{alg2}, on one hand, and control states $s_i$ of the subnets, on the other.}\st{
$L = {s_1, s_2, ...,  s_n}$;} We denote $\overline{ot}_i$ the set of open
transitions of the $i$-th
subnet; %% $s_i$ and 
``$-$'' means that the subnet is not involved.
The combination $\overline{comb}$, a set of n-tuples, is the
cartesian product:  
$\overline{comb} = (\inactive\cup \overline{ot}_1) \times (\inactive\cup \overline{ot}_2)\times \dots \times (\inactive\cup \overline{ot}_n)$\hl{, where, for each $i \in [1, n]$, $\overline{ot}_i$ is the set of open transitions obtained from the $i$-th subnet in line~3 of the algorithm}.

@@

\paragraph{Matching.}
The next step, for each synchronisation vector, and each possible
combination of behaviour of the subnets as generated by the \emph
{Combining} method, consists in building the combined predicate
and the corresponding open transition.
\hl{From a synchronization vector
$sv = \left((a'_i)^{i\in I} (b'_j)^{j\in J}\rightarrow v'\right) \in SV$
and its guard $G_k$; a tuple of open transitions
$C = ({ot}_i)^{i \in [1,n]} \in \overline{comb}$, such that, for each $i \in [1,n]$, either $ot_i = -$, or the result action of $ot_i$ is $a_i$; the hole behaviors $\mathit{Hole} = (b_j)^{j \in J}$; and a fresh variable $v$, representing the result action of the OT under construction, we build the predicate}
%
\begin{multline*}
  \mathit{Matching}(sv, C, \mathit{Hole}) = 
  \\
  (\forall i\in I, a_i=a'_i) \land
  (\forall i \not\in I, ot_i = -) \land
  (\forall j\in J, b_j=b'_j) \land
  (v=v') \land G_k.
\end{multline*}

\paragraph{Filtering.}
There may be several reasons why such a matching would fail:
\begin{itemize}
  \item if some subnet is marked as inactive in the vector, and the
    chosen combination has an active behaviour at this position,
  \item if some subnet action expression in the vector does not match (by
    pattern-matching) the corresponding \hl{action in $C$},
  \item or if the whole set of active subnet actions in the vector
    cannot be matched (by unification) with \hl{the tuple $C$}.
\end {itemize}

\st{Each of these possible filters is more precise than the previous
ones. But e}Even when unification succeeds, it is still possible that
the resulting predicate would be unsatisfiable, because of some
incompatibility involving the guards. In our algorithm, we choose to
apply only the simplest filter inside the \emph{Matching} method. All
the rest will be checked later, using the satisfiability check in the
SMT engine.

@@

In a pLTS, there may be several incoming transitions 
that assign potentially different values to a state variable.
To handle such cases, the algorithm manages, for each pLTS state, a
list of expressions collected from the assignments of each state
variable. %% $v$.
%% The list we used contains triples <$v$, $S$, $AssignRH$> 
%% where $v$ is the variable in pLTS, $S$ is its owner state and
%% $AssignRH$ is a list of expressions over other variables in the right
%% hand side of assignments of $v$.
% As a pragmatic extension to the formal definition, we also manage
% ``global variables'', defined in all states of the pLTS.
For a global state in the open automaton, getting the set of state
variables (which may be used in a transition) %% will simply be
\hl{is realised by taking
the union
of sets of} state variables of the individual pLTS states constituting this
global state (variables of different pLTS are distinct).
%There are some assignment on the variables in each of states of the pLTS (state variables), or just as global states of it (global variables). There is usually not only one assignment for a state variable, miss of them may drop the unsatisfiable results. In order to avoid such mistakes,

@@

Our matching/filtering strategy builds some transitions where
the predicates express incompatible constraints. Even if having an
unsatisfiable (symbolic) transition would not be incorrect,
\hl{we choose to minimize the open automaton (\ie its number of
transitions and states), by checking these predicates} for satisfiability.
In Fig.~\ref{schema:unsat-ot}, we show an 
unsatisfiable open transition 
from the open automaton of our running example. It shows the case where
the failure controller performs a ``$\PortFail$'' action, while the timer
executes a ``$\PortResume$''. The chosen synchronization
vector (SV0 from Fig.~\ref{schema:ArchFailure:pNet}) does not match
with these actions, \hl{since it expects $\NameTimer.\PortStart$. This mismatch is materialised} %% as shown
by the predicate
fragment ``$\mathit{resume}(\true)=\mathit{start}(b_2)$''.

@@

Checking satisfiability requires some symbolic computation
on the action expressions and the predicates, which
may depend on the specific theory of the action algebra datatypes. 
The ``Modulo Theory'' part of SMT solvers is important here, so that
the solver can use specific properties of each action algebra.

@@

We check the satisfiability of each open transitions
using the SMT solver Z3. More precisely, we translate
the OT predicate into an SMTlib assertion in the context provided by
the possible state variable assignments in the source state. In this
section, we describe the translation of the \noteSB{The algebra presentation only appears here for the first time after its definition in Sect.~\ref{section:pnets}}\hl{algebra presentation}, of
assignments, and of the predicates.

Our implementation submits satisfiability requests to
Z3 using its JAVA API. Here, for readability, we
show the Z3 code using its SMT-LIB input language.

Figure~\ref{schema:smt-lib} shows the translation of the
transition of Fig.~\ref{schema:unsat-ot}
in the SMTlib syntax.
It contains the declaration
of the BIP action algebra sorts 
and constructors, then the declaration of variables, and finally the
predicate to be checked, encoded as a set of assertions.
Here we display also the proof generated by Z3,
\noteSB{I am not sure I understand the entire message of this phrase}\hl{though in practice our tool only needs the brute information ``sat''
or ``unsat''. The result "unsat" in the output is just what we expected.}
% as appears in the ``unsat'' proof generated by Z3 in which ``\{= ?x66 ?x70\}'' is shown to be false (see Fig. \ref{schema:smt-lib})

@@

To build the input submitted to Z3 for each OT,
we translate the algebra presentation, the predicates and the
variable assignments (the Post part) into Z3 (Java-API) syntax.
% The $presentation$ and the predicates are translated separately.

\paragraph{Translation of action algebra presentation.}
In App.~\ref{appendix:mapsmt}, we formally define the condition of
well-formedness of an algebra presentation, and its translation into
SMTlib declarations (\texttt{declare-datatypes} and
\texttt{declare-fun}). It ensures that the generated code is correct,
and will not raise runtime errors in the SMT engine. Remark that the
\texttt{declare-datatypes} command comprises both the action
constructors from table \ref{table:BIPalgebra} and also the constant action
names from our example.

\paragraph{Translating assignments into predicate terms.}
State variable assignments that were collected during the open
automaton are used when generating the SMTlib asserts (see
App.~\ref{appendix:translating-assignments}). \noteSB{I am not an expert of Z3 or SMT-LIB, but the clause '\texttt{(= t (- t 1))}' looks very suspicious to me.}\hl{This appears in
line 10 of Fig.~\ref{schema:smt-lib}, which encodes possible values of
the variable $t$ of the timer.}

@@

\paragraph{Checking satisfiability.}

In App.~\ref{appendix:mapsmt} we formally define all steps of the translation of each
open transition, including:
%
\begin{itemize}
\item collect all variables in the transition, and declare them,
\item check well-formedness and correct typing of expressions,
\item translate the predicate into a conjunction of assertions,
\item if present in the source state, translate the state-variable
  assignments into a disjunctive assertion.
\end{itemize}
%
Here again, this translation ensures that no runtime error will occur
when running the SMT engine.
